DRAFT_Cleaning

### 2024-12-10 Draft_Cleaning.Rmd
cleaning up code, making final figures

```{r setup, include=FALSE}
library(readr)
library(tidyverse)
library(lubridate)
library(caret)
library(ggfortify)
library(ggformula)
library(DiagrammeR)
```

Read in the data
```{r}
dredge_data<-read_csv("DredgeData.csv", show_col_types = FALSE)
survey_data<-read_csv("SurveyData.csv", show_col_types = FALSE)
gage_data<-read_csv("UMR_IWW_1999_2021.csv", show_col_types = FALSE)

### The gage data has multiple records across different gages per day, we need to
### combine the rows by date so that there's only one record for date across all gages.
### Since we're going to be doing a lot of work with the date column, lets format it as 
### a date object and arrange by DATE_START and remove the old Date column
gage_data<- gage_data |>
  group_by(Date)|>
   summarize(across(everything(),~first(na.omit(.)), .names = "{.col}"))|>
  mutate(DATE_START= as_date(dmy(Date)))|>
    arrange(DATE_START)|>
  select(-Date)

@@

Now lets clean the survey data. I only want conditional survey records between
1999 to 2000, these are going to act as my background data. I want to combine
these records with my dredge data. I'm also going to filter by river, remove
pools that are not of interest, clean up the SURVEYJOBIDPK, and create some 
columns to support my join. Dates are also formatted as date objects and some
columns are renamed to make it easier to join my surveys to my dredge data.
Finally, select the columns we need for our analysis to join to our dredging 
data. 
```{r}
survey_data<- survey_data |>
  filter(SURVEYTYPE == "conditional",
         SURVEY_YEAR >= 1999 & SURVEY_YEAR <= 2022,
          RIVER == "Mississippi_River" | RIVER == "Illinois_Waterway",
         !POOL %in% c("LP","CS"))|>
  mutate(VOLUMEDREDGED = 0,
         DREDGINGPURPOSE = "None",
         DREDGE_TYPE = "None",
         LABOR_TYPE = LABOR_SOURCE,
         EXECYEAR = SURVEY_YEAR,
         FEATURENAME = SDSFEATURENAME,
         DATE_START = as.character(DATE_START),
         DATE_START = as_date(DATE_START)
         )|>
  select(EXECYEAR,DATE_START,DREDGINGPURPOSE,VOLUMEDREDGED, RIVER,POOL, DREDGE_TYPE,
         UP_RIV_MIL,DN_RIV_MIL)
```

@@

@@

gage_survey_filt <- gage_survey |>
  filter(complete.cases(gage_survey))

gage_survey_full <- gage_survey_filt|>  
  left_join(gage_data, by = c("lead_7" = "DATE_START"), suffix = c("", "_7"))|>
  left_join(gage_data, by = c("lead_14" = "DATE_START"), suffix = c("","_14"))

Nas<- gage_survey_full|>
  select(-lead_7,-lead_14)|>
  filter(if_any(everything(),is.na))

@@

Create factored season variable that could be used in analysis
```{r}
date_to_season <- function(date) {
  month <- as.numeric(format(date, "%m"))
  day <- as.numeric(format(date, "%d"))
  # Define the seasons based on date ranges
  if ((month == 3 && day >= 20) || (month >= 4 && month <= 6) || (month == 6 && day <= 20)) {
    return("Spring")
  } else if ((month == 6 && day >= 21) || (month >= 7 && month <= 9) || (month == 9 && day <= 21)) {
    return("Summer")
  } else if ((month == 9 && day >= 22) || (month >= 10 && month <= 12) || (month == 12 && day <= 20)) {
    return("Fall")
  } else {
    return("Winter")
  }
}

gage_survey_cleaned <- gage_survey_cleaned |>
  mutate("Season" = sapply(DATE_START,date_to_season))

```{r}
gage_survey_PCA <- gage_survey_cleaned |>
select(where(is.numeric), -EXECYEAR)

dredge_numeric<-gage_survey_cleaned|>
   select(where(is.numeric), -EXECYEAR)
correlations <-cor(dredge_numeric, use = "pairwise.complete.obs")
corrplot::corrplot(correlations)

@@

biplot(dredge_PCA)

dredge <-gage_survey_cleaned |>
  filter(DREDGINGPURPOSE != "None")
  hist(dredge$VOLUMEDREDGED)

  median(dredge$VOLUMEDREDGED)

hist(gage_survey_cleaned$VOLUMEDREDGED)

@@

autoplot(dredge_PCA,data = gage_survey_cleaned, 
         color = 'SEASON',loadings = TRUE,
         loadings.colour = 'black',
         loadings.label = TRUE, loadings.label.size = 4, 
         loadings.label.colour = "blue")

variable_contributions<- dredge_PCA$rotation
head(variable_contributions)

set.seed(12)
ctrl = trainControl(method = "cv", number = 5)

@@

# data.frame(variable_contributions) |>
#   mutate(RIVER = gage_survey_cleaned$RIVER)
#   gf_point(PC2 ~ PC1)

@@

data_in_PC_space = data.frame(data_in_PC_space,
                              River = gage_survey_cleaned$RIVER,
                              Pool = gage_survey_cleaned$POOL,
                              DREDGE_TYPE = gage_survey_cleaned$DREDGINGPURPOSE,
                              Season = gage_survey_cleaned$Season)

@@

var_df|>
  gf_point(cum_var ~ PC) |>
  gf_line(cum_var ~ PC)

pca_scores <-dredge_PCA$x
pca_loadings <- dredge_PCA$rotation

###xgBoost
```{r}
library(xgboost)
gage_xgb <- gage_survey_cleaned |>
  select(-lead_7, -lead_14,-DATE_START,
         -SURVEYJOBIDPK,-EXECYEAR,-DREDGE_TYPE,
         -DREDGINGPURPOSE)|>
    mutate_if(is.character, factor)

@@

import_matrix = xgb.importance(model = fit_dredge$finalModel)
import_matrix
xgb.plot.importance(import_matrix)
```
###Cross Validation of Model
```{r}
set.seed(123)
n = nrow(gage_xgb)
n_outer_folds = 5

@@

### Cleaning the data 
```{r}
combined_data <- bind_rows(survey_Data, dredge_data_cleaning)

@@

Illinois_River <- gage_survey |>
  select(SURVEYJOBIDPK,EXECYEAR,DATE_START,DREDGINGPURPOSE,VOLUMEDREDGED,RIVER,
         POOL, DREDGE_TYPE,UP_RIV_MIL, DN_RIV_MIL,lead_7,lead_14,BEAI2,HAVI2,
         HNYI2,IL03, IL03_Tail, IL04, IL04_Tail, IL05, IL05_Tail,IL06, IL06_Tail,
         IL07, IL07_Tail, IL08, IL08_Tail,KNGI2, MORI2, PIAI2) |>
  filter(RIVER == "Illinois_Waterway")

Miss_River <- gage_survey |>
  select(SURVEYJOBIDPK,EXECYEAR,DATE_START,DREDGINGPURPOSE,VOLUMEDREDGED,RIVER,
         POOL, DREDGE_TYPE,UP_RIV_MIL, DN_RIV_MIL,lead_7,lead_14,BRLI4,CMMI4, DBQI4,
         FAII4, KHBI2,MI11,MI11_Tail,MI12,MI12_Tail,MI13,MI13_Tail,MI14,MI14_Tail,
         MI15,MI15_Tail,MI16,MI16_Tail,MI17,MI17_Tail,MI18,MI18_Tail,MI19,MI19_Tail,
         MI20,MI20_Tail,MI21,MI21_Tail,MI22,MI22_Tail, MUSI4,UINI2) |>
  filter(RIVER == "Mississippi_River")

Miss_full <- Miss_River|>  
  left_join(gage_data, by = c("lead_7" = "DATE_START"), suffix = c("", "_7"))|>
  left_join(gage_data, by = c("lead_14" = "DATE_START"), suffix = c("","_14"))

IWW_full <- Illinois_River|>  
  left_join(gage_data, by = c("lead_7" = "DATE_START"), suffix = c("", "_7"))|>
  left_join(gage_data, by = c("lead_14" = "DATE_START"), suffix = c("","_14"))

@@

## PCA
```{r}

@@

## xgboost split by river
#Miss
```{r}
Miss_xgb <-Miss_Cleaned |>
  select(-lead_7, -lead_14,-DATE_START,
         -SURVEYJOBIDPK,-EXECYEAR,-DREDGE_TYPE,
         -DREDGINGPURPOSE, -RIVER)|>
    mutate_if(is.character, factor)

@@

#IWW
```{r}
IWW_xgb <-IWW_Cleaned |>
  select(-lead_7, -lead_14,-DATE_START,
         -SURVEYJOBIDPK,-EXECYEAR,-DREDGE_TYPE,
         -DREDGINGPURPOSE, -RIVER)|>
    mutate_if(is.character, factor)

>>>>

```{r setup, include=FALSE}
library(readr)
library(tidyverse)
library(lubridate)
library(caret)
library(ggfortify)
library(ggformula)
library(DiagrammeR)
library(gridExtra)

### Key Functions 
# Function to plot PCA
plot_pca <- function(PCAvalues, top_loadings, PC1_explained, PC2_explained, grouping_var) {

  p <- ggplot(PCAvalues, aes(x = "PC1", y = "PC2", colour = grouping_var)) +
    geom_point(size = 2) +  # Scatter plot
    geom_segment(data = top_loadings, aes(x = 0, y = 0, xend = PC1 * 75, yend = PC2 * 75),
                 arrow = arrow(length = unit(1/2, "picas")), color = "black") +  # Vectors for top loadings
    annotate("text", x = top_loadings$PC1 * 80, y = top_loadings$PC2 * 80,
             label = top_loadings$Variables, hjust = 0.2, vjust = 0.4) +  # Labels for top loadings
    theme_minimal() +
    labs(title = paste("PCA Plot with Top 10 Loadings for PC1 and PC2 - Grouped by", grouping_var),
         x = paste("PC1 (", round(PC1_explained, 1), "%)", sep = ""),
         y = paste("PC2 (", round(PC2_explained, 1), "%)", sep = ""))

  return(p)
}

#Function to calculate season from date 
date_to_season <- function(date) {
  month <- as.numeric(format(date, "%m"))
  day <- as.numeric(format(date, "%d"))
  # Define the seasons based on date ranges
  if ((month == 3 && day >= 20) || (month >= 4 && month <= 6) || (month == 6 && day <= 20)) {
    return("Spring")
  } else if ((month == 6 && day >= 21) || (month >= 7 && month <= 9) || (month == 9 && day <= 21)) {
    return("Summer")
  } else if ((month == 9 && day >= 22) || (month >= 10 && month <= 12) || (month == 12 && day <= 20)) {
    return("Fall")
  } else {
    return("Winter")
  }
}

### Read in the data
```{r}
dredge_data<-read_csv("DredgeData.csv", show_col_types = FALSE)
survey_data<-read_csv("SurveyData.csv", show_col_types = FALSE)
gage_data<-read_csv("UMR_IWW_1999_2021.csv", show_col_types = FALSE)
```

### Clean the gage data 
The gage data has multiple records across different gages per day, we need to
combine the rows by date so that there's only one record for date across all gages.
Since we're going to be doing a lot of work with the date column, lets format it as 
a date object and arrange by DATE_START and remove the old Date column
```{r}
gage_data<- gage_data |>
  group_by(Date)|>
   summarize(across(everything(),~first(na.omit(.)), .names = "{.col}"))|>
  mutate(DATE_START= as_date(dmy(Date)))|>
    arrange(DATE_START)|>
  select(-Date)

@@

### Cleaning the data
Now lets clean the survey data. I only want conditional survey records between
1999 to 2000, these are going to act as my background data. I want to combine
these records with my dredge data. I'm also going to filter by river, remove
pools that are not of interest, and create some 
columns to support my join. Dates are also formatted as date objects and some
columns are renamed to make it easier to join my surveys to my dredge data.
Finally, select the columns we need for our analysis to join to our dredging 
data. 
```{r}
survey_data<- survey_data |>
  filter(SURVEYTYPE == "conditional",
         SURVEY_YEAR >= 1999 & SURVEY_YEAR <= 2022,
          RIVER == "Mississippi_River" | RIVER == "Illinois_Waterway",
         !POOL %in% c("LP","CS"))|>
  mutate(VOLUMEDREDGED = 0,
         DREDGINGPURPOSE = "None",
         DREDGE_TYPE = "None",
         LABOR_TYPE = LABOR_SOURCE,
         EXECYEAR = SURVEY_YEAR,
         FEATURENAME = SDSFEATURENAME,
         DATE_START = as.character(DATE_START),
         DATE_START = as_date(DATE_START)
         )|>
  select(EXECYEAR,DATE_START,DREDGINGPURPOSE,VOLUMEDREDGED, RIVER,POOL, DREDGE_TYPE,
         UP_RIV_MIL,DN_RIV_MIL)
```

@@

combined_data <-combined_data |>
     filter(DATE_START >= as.Date("1999-04-01") &
         DATE_START <= as.Date("2021-09-30")) |>
          mutate( DREDGINGPURPOSE = case_when(DREDGINGPURPOSE == "None" ~ "No Dredging",
                                      DREDGINGPURPOSE == "channel" ~ "Dredged"))

combined_data <-combined_data |>
  mutate(lead_7 = (DATE_START - 7),
         lead_14 = (DATE_START - 14))
```

Check for NAs for PCA analysis. 
```{r}
Nas<- combined_data|>
  filter(if_any(everything(),is.na))

@@

gage_survey_filt <- gage_survey |>
  filter(complete.cases(gage_survey))
```

Join the gage data by the 7 day prior observations and by the 14 day prior observations
```{r}
gage_survey_full <- gage_survey_filt|>  
  left_join(gage_data, by = c("lead_7" = "DATE_START"), suffix = c("", "_7"))|>
  left_join(gage_data, by = c("lead_14" = "DATE_START"), suffix = c("","_14"))
```

Check again for NAs, remove any remaining NA rows. 
```{r}
Nas<- gage_survey_full|>
  select(-lead_7,-lead_14)|>
  filter(if_any(everything(),is.na))

@@

Create factored season variable that could be used in analysis
```{r}
gage_survey_cleaned <- gage_survey_cleaned |>
  mutate("Season" = sapply(DATE_START,date_to_season))

```
Now on to analysis!

To first explore the variables, we will use PCA.
### PCA
PCA using prcomp
```{r}
gage_survey_PCA <- gage_survey_cleaned |>
select(where(is.numeric), -EXECYEAR)

@@

@@

a<-autoplot(dredge_PCA,data = gage_survey_cleaned, 
         color = 'Season',loadings = FALSE)+
  ggtitle("Season")

b<-autoplot(dredge_PCA,data = gage_survey_cleaned, 
         color = 'POOL',loadings= FALSE)+
  ggtitle("River Pool")+
    guides(color = guide_legend(ncol = 2))

c<-autoplot(dredge_PCA,data = gage_survey_cleaned, 
         color = 'RIVER',loadings = FALSE)+
  ggtitle("River")

d<-autoplot(dredge_PCA,data = gage_survey_cleaned, 
         color = 'DREDGINGPURPOSE',loadings = FALSE)+
        ggtitle("Dredging Purpose")+
  labs(color = "Dredging?")

ggsave("Initial_Full_PCA.png",grid.arrange(a,b,c,d,ncol = 2),width = 10, height = 8)
variable_contributions<- dredge_PCA$rotation
head(variable_contributions)
```

PCA using Caret
```{r}
set.seed(12)
ctrl = trainControl(method = "cv", number = 5)

@@

@@

data_in_PC_space = data.frame(data_in_PC_space,
                              River = gage_survey_cleaned$RIVER,
                              Pool = gage_survey_cleaned$POOL,
                              DREDGE_TYPE = gage_survey_cleaned$DREDGINGPURPOSE,
                              Season = gage_survey_cleaned$Season,
                              Dredge_Volume = gage_survey_cleaned$VOLUMEDREDGED)

@@

var_df|>
  gf_point(cum_var ~ PC) |>
  gf_line(cum_var ~ PC)
```

###xgBoost
```{r}
library(xgboost)
gage_xgb <- gage_survey_cleaned |>
  select(-lead_7, -lead_14,-DATE_START,
         -EXECYEAR,-DREDGE_TYPE,
         -DREDGINGPURPOSE)|>
    mutate_if(is.character, factor)

@@

import_matrix = xgb.importance(model = fit_dredge$finalModel)
import_matrix
xgb.plot.importance(import_matrix)

```
###Cross Validation of Model
```{r}
set.seed(123)
n = nrow(gage_xgb)
n_outer_folds = 5

@@

### Cleaning the data 
```{r}
combined_data <- bind_rows(survey_data, dredge_data)

@@

gage_IWW <- gage_data |>
  select(DATE_START,BEAI2,HAVI2,
         HNYI2,IL03, IL03_Tail, IL04, IL04_Tail, IL05, IL05_Tail,IL06, IL06_Tail,
         IL07, IL07_Tail, IL08, IL08_Tail,KNGI2, MORI2, PIAI2)

gage_Miss <- gage_data |>
  select(DATE_START,BRLI4,CMMI4, DBQI4,
         FAII4, KHBI2,MI11,MI11_Tail,MI12,MI12_Tail,MI13,MI13_Tail,MI14,MI14_Tail,
         MI15,MI15_Tail,MI16,MI16_Tail,MI17,MI17_Tail,MI18,MI18_Tail,MI19,MI19_Tail,
         MI20,MI20_Tail,MI21,MI21_Tail,MI22,MI22_Tail, MUSI4,UINI2)

Illinois_River <- gage_survey |>
  select(EXECYEAR,DATE_START,DREDGINGPURPOSE,VOLUMEDREDGED,RIVER,
         POOL, DREDGE_TYPE,UP_RIV_MIL, DN_RIV_MIL,lead_7,lead_14,BEAI2,HAVI2,
         HNYI2,IL03, IL03_Tail, IL04, IL04_Tail, IL05, IL05_Tail,IL06, IL06_Tail,
         IL07, IL07_Tail, IL08, IL08_Tail,KNGI2, MORI2, PIAI2) |>
  filter(RIVER == "Illinois_Waterway")

Miss_River <- gage_survey |>
  select(EXECYEAR,DATE_START,DREDGINGPURPOSE,VOLUMEDREDGED,RIVER,
         POOL, DREDGE_TYPE,UP_RIV_MIL, DN_RIV_MIL,lead_7,lead_14,BRLI4,CMMI4, DBQI4,
         FAII4, KHBI2,MI11,MI11_Tail,MI12,MI12_Tail,MI13,MI13_Tail,MI14,MI14_Tail,
         MI15,MI15_Tail,MI16,MI16_Tail,MI17,MI17_Tail,MI18,MI18_Tail,MI19,MI19_Tail,
         MI20,MI20_Tail,MI21,MI21_Tail,MI22,MI22_Tail, MUSI4,UINI2) |>
  filter(RIVER == "Mississippi_River")

Miss_full <- Miss_River|>  
  left_join(gage_Miss, by = c("lead_7" = "DATE_START"), suffix = c("", "_7"))|>
  left_join(gage_Miss, by = c("lead_14" = "DATE_START"), suffix = c("","_14"))

IWW_full <- Illinois_River|>  
  left_join(gage_IWW, by = c("lead_7" = "DATE_START"), suffix = c("", "_7"))|>
  left_join(gage_IWW, by = c("lead_14" = "DATE_START"), suffix = c("","_14"))

@@

## IWW PCA
```{r}
IWW_PCA <- IWW_Cleaned |>
select(where(is.numeric), -EXECYEAR)

IWW_prcomp <- IWW_PCA|>
  select(-VOLUMEDREDGED)

IWW_PCA = prcomp(IWW_prcomp, scale = T)

biplot(IWW_PCA)

plot(IWW_PCA,  type="l")

autoplot(IWW_PCA,data = IWW_Cleaned, 
         color = 'Season',loadings = TRUE,
         loadings.colour = 'black',
         loadings.label = TRUE, loadings.label.size = 4, 
         loadings.label.colour = "blue")

PCAvalues <- data.frame(Pool = IWW_Cleaned$POOL,
                        Dredge_Type = IWW_Cleaned$DREDGE_TYPE,
                        Season = IWW_Cleaned$Season,
                        Dredge_Occurance = IWW_Cleaned$DREDGINGPURPOSE,
                        IWW_PCA$x)

explained_variance <- (IWW_PCA$sdev^2) / sum(IWW_PCA$sdev^2) * 100

# Extract PC1 and PC2 explained variance
PC1_explained <- explained_variance[1]
PC2_explained <- explained_variance[2]

# Extract loadings of the variables
PCAloadings <- data.frame(Variables = rownames(IWW_PCA$rotation), IWW_PCA$rotation)

PCAloadings_PC1 <- PCAloadings|>
  arrange(desc(abs(PC1))) |>
  slice_head(n=10)

PCAloadings_PC2 <- PCAloadings|>
  arrange(desc(abs(PC2))) |>
  slice_head(n=10)

top_loadings <- bind_rows(PCAloadings_PC1, PCAloadings_PC2)

# Plot PCA with top 10 loadings
ggplot(PCAvalues, aes(x = PC1, y = PC2, colour = Season)) +
  geom_point(size = 2) +
  geom_segment(data = top_loadings, aes(x = 0, y = 0, xend = PC1 * 75, yend = PC2 * 75),
               arrow = arrow(length = unit(1/2, "picas")), color = "black") +  
  annotate("text", x = top_loadings$PC1 * 80, y = top_loadings$PC2 * 80,
           label = top_loadings$Variables, hjust = 0.2, vjust = 0.4) +  
  theme_minimal() +
  labs(title = "PCA Plot with Top 10 Loadings for PC1 and PC2",
      x = paste("PC1 (", round(PC1_explained, 1), "%)", sep = ""),
       y = paste("PC2 (", round(PC2_explained, 1), "%)", sep = ""))

ggplot(PCAvalues, aes(x = PC1, y = PC2, colour = Pool)) +
  geom_point(size = 2) +
  geom_segment(data = top_loadings, aes(x = 0, y = 0, xend = PC1 * 75, yend = PC2 * 75),
               arrow = arrow(length = unit(1/2, "picas")), color = "black") +  
  annotate("text", x = top_loadings$PC1 * 80, y = top_loadings$PC2 * 80,
           label = top_loadings$Variables, hjust = 0.2, vjust = 0.4) +  
  theme_minimal() +
  labs(title = "PCA Plot with Top 10 Loadings for PC1 and PC2",
      x = paste("PC1 (", round(PC1_explained, 1), "%)", sep = ""),
       y = paste("PC2 (", round(PC2_explained, 1), "%)", sep = ""))

ggplot(PCAvalues, aes(x = PC1, y = PC2, colour = Dredge_Type)) +
  geom_point(size = 2) + 
  geom_segment(data = top_loadings, aes(x = 0, y = 0, xend = PC1 * 75, yend = PC2 * 75),
               arrow = arrow(length = unit(1/2, "picas")), color = "black") + 
  annotate("text", x = top_loadings$PC1 * 80, y = top_loadings$PC2 * 80,
           label = top_loadings$Variables, hjust = 0.2, vjust = 0.4) +
  theme_minimal() +
  labs(title = "PCA Plot with Top 10 Loadings for PC1 and PC2",
      x = paste("PC1 (", round(PC1_explained, 1), "%)", sep = ""),
       y = paste("PC2 (", round(PC2_explained, 1), "%)", sep = ""))

ggplot(PCAvalues, aes(x = PC1, y = PC2, colour = Dredge_Occurance)) +
  geom_point(size = 2) + 
  geom_segment(data = top_loadings, aes(x = 0, y = 0, xend = PC1 * 75, yend = PC2 * 75),
               arrow = arrow(length = unit(1/2, "picas")), color = "black") +
  annotate("text", x = top_loadings$PC1 * 80, y = top_loadings$PC2 * 80,
           label = top_loadings$Variables, hjust = 0.2, vjust = 0.4) + 
  theme_minimal() +
  labs(title = "PCA Plot with Top 10 Loadings for PC1 and PC2",
      x = paste("PC1 (", round(PC1_explained, 1), "%)", sep = ""),
       y = paste("PC2 (", round(PC2_explained, 1), "%)", sep = ""))
```

## Miss PCA 
```{r}
Miss_PCA <- Miss_Cleaned |>
select(where(is.numeric), -EXECYEAR)

Miss_prcomp <- Miss_PCA|>
  select(-VOLUMEDREDGED)

Miss_PCA = prcomp(Miss_prcomp, scale = T)

biplot(Miss_PCA)
biplot(Miss_PCA, showLoadings = TRUE,)

plot(Miss_PCA,  type="l")

PCAvalues <- data.frame(Pool = Miss_Cleaned$POOL,
                        Dredge_Type = Miss_Cleaned$DREDGE_TYPE,
                        Season = Miss_Cleaned$Season,
                        Dredge_Occurance = Miss_Cleaned$DREDGINGPURPOSE,
                        Miss_PCA$x)

explained_variance <- (Miss_PCA$sdev^2) / sum(Miss_PCA$sdev^2) * 100

# Extract PC1 and PC2 explained variance
PC1_explained <- explained_variance[1]
PC2_explained <- explained_variance[2]

# Extract loadings of the variables
PCAloadings <- data.frame(Variables = rownames(Miss_PCA$rotation), Miss_PCA$rotation)

PCAloadings_PC1 <- PCAloadings|>
  arrange(desc(abs(PC1))) |>
  slice_head(n=10)

PCAloadings_PC2 <- PCAloadings|>
  arrange(desc(abs(PC2))) |>
  slice_head(n=10)

top_loadings <- bind_rows(PCAloadings_PC1, PCAloadings_PC2)

# Plot PCA with top 10 loadings
ggplot(PCAvalues, aes(x = PC1, y = PC2, colour = Season)) +
  geom_point(size = 2) +  # Scatter plot
  geom_segment(data = top_loadings, aes(x = 0, y = 0, xend = PC1 * 75, yend = PC2 * 75),
               arrow = arrow(length = unit(1/2, "picas")), color = "black") +  # Vectors for top loadings
  annotate("text", x = top_loadings$PC1 * 80, y = top_loadings$PC2 * 80,
           label = top_loadings$Variables, hjust = 0.2, vjust = 0.4) +  # Labels for top loadings
  theme_minimal() +
  labs(title = "PCA Plot with Top 10 Loadings for PC1 and PC2",
      x = paste("PC1 (", round(PC1_explained, 1), "%)", sep = ""),
       y = paste("PC2 (", round(PC2_explained, 1), "%)", sep = ""))

ggplot(PCAvalues, aes(x = PC1, y = PC2, colour = Pool)) +
  geom_point(size = 2) +  # Scatter plot
  geom_segment(data = top_loadings, aes(x = 0, y = 0, xend = PC1 * 75, yend = PC2 * 75),
               arrow = arrow(length = unit(1/2, "picas")), color = "black") +  # Vectors for top loadings
  annotate("text", x = top_loadings$PC1 * 80, y = top_loadings$PC2 * 80,
           label = top_loadings$Variables, hjust = 0.2, vjust = 0.4) +  # Labels for top loadings
  theme_minimal() +
  labs(title = "PCA Plot with Top 10 Loadings for PC1 and PC2",
      x = paste("PC1 (", round(PC1_explained, 1), "%)", sep = ""),
       y = paste("PC2 (", round(PC2_explained, 1), "%)", sep = ""))

ggplot(PCAvalues, aes(x = PC1, y = PC2, colour = Dredge_Type)) +
  geom_point(size = 2) +  # Scatter plot
  geom_segment(data = top_loadings, aes(x = 0, y = 0, xend = PC1 * 75, yend = PC2 * 75),
               arrow = arrow(length = unit(1/2, "picas")), color = "black") +  # Vectors for top loadings
  annotate("text", x = top_loadings$PC1 * 80, y = top_loadings$PC2 * 80,
           label = top_loadings$Variables, hjust = 0.2, vjust = 0.4) +  # Labels for top loadings
  theme_minimal() +
  labs(title = "PCA Plot with Top 10 Loadings for PC1 and PC2",
      x = paste("PC1 (", round(PC1_explained, 1), "%)", sep = ""),
       y = paste("PC2 (", round(PC2_explained, 1), "%)", sep = ""))

@@

## xgboost split by river
#Miss
```{r}
Miss_xgb <-Miss_Cleaned |>
  select(-lead_7, -lead_14,-DATE_START,
         -EXECYEAR,-DREDGE_TYPE,
         -DREDGINGPURPOSE, -RIVER)|>
    mutate_if(is.character, factor)

@@

#IWW
```{r}
IWW_xgb <-IWW_Cleaned |>
  select(-lead_7, -lead_14,-DATE_START,
         -EXECYEAR,-DREDGE_TYPE,
         -DREDGINGPURPOSE, -RIVER)|>
    mutate_if(is.character, factor)
